Isomorphisms as Groupoids: \\ Homotopy Type Theory with Fractional Types

### 2013-10-16 f.tex
groupoids

\newcommand{\union}{\cup}
\newtheorem{definition}{Definition}
\newcommand{\refl}{\mathit{refl}}
\renewcommand{\path}{\leadsto}
\renewcommand{\circ}{\circlearrowleft}
\newcommand{\alt}{~|~}
\newcommand{\dt}[1]{\llbracket #1 \rrbracket}
\newcommand{\leftv}[1]{\textsf{left}~#1}
\newcommand{\rightv}[1]{\textsf{right}~#1}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\wedgesum}{\vee}
\newcommand{\smashproduct}{\wedge}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\proves}{\vdash}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}

@@

@@

The language presented above, at the type level, models a \emph{commutative
  semiring} (occasionally called a \emph{commutative rig}) where we replace
equality by isomorphism.  Semantically, $\Pi$ models a \emph{bimonoidal
  category} whose simplest example is the category of finite sets and
bijections. In that interpretation, each value type denotes a finite set of a
size calculated by viewing the types as natural numbers and each combinator
$c : b_1 \iso b_2$ denotes a bijection between the sets denoted by $b_1$ and
$b_2$. Instead of presenting such a semantics, we present a slightly more
involved semantics using discrete groupoids. In that semantics, the
denotation of each type will consist of (i) a pointed set, and (ii) a trivial
equivalence relation relating each element to itself by a \emph{path}. The
pointed sets are constructed by adjoining, to each ``regular'' set, an
element $\bot$ with a non-trivial path to itself which, topologically,
corresponds to the \emph{circle}. Significantly, using the Euler
characteristic as a generalized notion of cardinality, a set containing $n$
discrete elements (with only trivial reflexivity paths) and a circle has
cardinality $n$ and not $n+1$.\footnote{Generally speaking, this generalized
  cardinality based on the Euler characteristic counts the number of
  ``connected components'' of a set. Consider a set with $n$ points and
  various paths connecting them. We start by counting all the points giving
  us cardinality $n$. Then for every path, we decrease the cardinality by one
  to avoid double-counting connected elements. Then we add one again for
  every path that forms a loop because the path forming the loop did not
  decrease the number of connected components. It is easy to see that a point
  with a path that connects it to itself would have generalized cardinality
  0. Moreover the generalized cardinality of the disjoint union of two sets
  is their sum and the generalized cardinality of the product of two sets is
  their product.}

\begin{definition}[Denotation of types]
Each type $b$ is mapped to a groupoid $(\dt{b}, \bot_b, \mathit{paths}_b)$
where~$\dt{b}$ is a pointed set with distinguished point $\bot_b$ and
$\mathit{paths}_b$ is a collection of paths, $v_1 \path v_2$,
between elements $v_1$ and $v_2$ of $\dt{b}$. (We omit the trivial paths for
reflexivity and only include the non-trivial ``generating'' paths as
suggested by the notation of \emph{higher-order inductive types}.)
\[\begin{array}{rcl@{\qquad}l}
\dt{0} &=& (\{ \bot_0 \}, \bot_0, \{ \bot_0 \path \bot_0 \}) &
           (\mbox{circle}) \\
\dt{1} &=& (\{ \bullet, \bot_1 \}, \bot_1, \{ \bot_1 \path \bot_1 \}) &\\
\dt{(b_1 + b_2)} &=& \dt{b_1} \wedgesum \dt{b_2} & 
           (\mbox{wedge~sum}) \\
\dt{(b_1 * b_2)} &=& \dt{b_1} \smashproduct \dt{b_2} & 
           (\mbox{smash~product})
\end{array}\]
The empty type is mapped to the topological circle which is a point that is
connected to itself. The unit type is mapped to a pointed singleton with a
distinguished element connected to itself with a non-trivial loop. The
operation $\wedgesum$ is the \emph{wedge sum} which corresponds to the
disjoint union with an additional path $\leftv{\bot_{b_1}} \path
\rightv{\bot_{b_2}}$ identifying the two distinguished elements of the
component spaces. The new cluster is the distinguished element of the
composite space. The operation $\smashproduct$ is the \emph{smash product}
which is the regular set product with additional paths connecting
$(x,\bot_{b_2})$ and $(\bot_{b_1},y)$ for all $x \in \dt{b_1}$ and $y \in
\dt{b_2}$. The new cluster of identified distinguished elements is the new
distinguished element of the composite space.
\end{definition}

We quickly check that the desired isomorphisms hold at the level of types,
e.g., the isomorphism $b * 0 \iso 0$ requires that the smash product of any
groupoid with the circle be indistinguishable from the circle.

@@

>>>>

\newcommand{\union}{\cup}
\newcommand{\transport}[2]{\overline{#1}(#2)}
\newtheorem{definition}{Definition}
\newcommand{\refl}{\mathit{refl}}
\renewcommand{\path}{\leadsto}
%% \renewcommand{\circ}{\circlearrowleft}
\newcommand{\alt}{~|~}
\newcommand{\dt}[1]{\llbracket #1 \rrbracket}
\newcommand{\leftv}[1]{\textsf{left}~#1}
\newcommand{\rightv}[1]{\textsf{right}~#1}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\wedgesum}{\vee}
\newcommand{\smashproduct}{\wedge}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\proves}{\vdash}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}

@@

%%%%%%%%%%%%%%%%%
\subsection{Types as Groupoids}

The language presented above, at the type level, models a \emph{commutative
  semiring} (occasionally called a \emph{commutative rig}) where we replace
equality by isomorphism.  Semantically, $\Pi$ models a \emph{bimonoidal
  category} whose simplest example is the category of finite sets and
bijections. In that interpretation, each value type denotes a finite set of a
size calculated by viewing the types as natural numbers and each combinator
$c : b_1 \iso b_2$ denotes a bijection between the sets denoted by $b_1$ and
$b_2$. Instead of presenting such a semantics, we present a more involved
semantics using groupoids.

\begin{definition}[Path]
A \emph{path} $p : x \path y$ between two points $x$ and $y$ in some space
$A$ is a directed edge $p$ from $x$ to $y$. The notation $\transport{p}{x}$
denotes the action of \emph{transporting} the point $x$ to the point~$y$
along $p$.
\end{definition}

\begin{definition}[Disjoint union of paths]
Given two sets of paths $P$ and $Q$, the disjoint union $P \uplus^p Q$ is
defined as:
\[ 
  \{ \leftv{p} : \leftv{x} \path \leftv{y} ~|~ (p : x \path y) \in P \}
  \union
  \{ \rightv{q} : \leftv{x} \path \leftv{y} ~|~ (q : x \path y) \in Q \}
\]
\end{definition}
We will sometimes write this as $\leftv{P} \union \rightv{Q}$.

\begin{definition}[Product of paths]
Given two sets of paths $P$ and $Q$, the product $P \times^p Q$ is
defined as:
\[ 
\{ (p,q) : (x,z) \path (y,w) ~|~ (p : x \path y) \in P, (q : z \path w) \in Q \}
\]
\end{definition}
We will sometimes write this as $(P,Q)$.

\begin{definition}[Groupoid]
A \emph{groupoid} $G = (G_0, \mathit{paths}_{G_0})$ is a set $G_0$ together
with a collection of paths $x \path y$, between elements $x$ and $y$ of $G_0$
that satisfies the following conditions:
\begin{itemize}
\item for every $x \in G_0$, there is a path $\refl_x : x \path x$,
\item every path $p : x \path y$, there is an inverse path $! p : y \path x$, 
\item for every paths $p : x \path y$ and $q : y \path z$ there is a path $p
  \circ q : x \path z$, and
\item for every paths $p : x \path y$, $q : y \path z$, and $r : z \path w$,
  we have:
  \begin{itemize}
  \item $\refl_x \circ p = p$;
  \item $p \circ \refl_y = p$;
  \item $(p \circ q) \circ r = (p \circ (q \circ r))$;
  \item $p \,\circ\, !p = \refl_x$;
  \item $!p \,\circ\, p = \refl_y$.
  \end{itemize}
\end{itemize}
\end{definition}
When introducing a groupoid, we only explicitly mention a collection of
non-trivial ``generating'' paths. The equivalence closure under $\refl$, $!$,
and $\circ$ with the appropriate conditions is implicitly
assumed.\footnote{For higher groupoids, the five equivalences in the last
  condition will be witnessed by paths between paths and not by the
  extensional equivalence relation $=$.}

\begin{definition}[Pointed groupoid]
A pointed groupoid $G_{\bullet} = (G_0, \bot_{G_0}, \mathit{paths}_{G_0})$ is
a groupoid $(G_0, \mathit{paths}_{G_0})$ with a distinguished element
$\bot_{G_0} \in G_0$.
\end{definition}

\begin{definition}[Wedge sum]
The \emph{wedge sum} of two pointed groupoids $G_\bullet$ and $H_\bullet$,
written $G_\bullet \wedgesum H_\bullet$ is defined as follows:
\[\begin{array}{l}
(G_0, \bot_{G_0}, \mathit{paths}_{G_0}) \wedgesum 
(H_0, \bot_{H_0}, \mathit{paths}_{H_0}) = \\
((G_0 \uplus H_0) \union \{ \bot \},
 \bot,
 (\mathit{paths}_{G_0} \uplus^p \mathit{paths}_{H_0}) \union
 \{ L: \leftv{\bot_{G_0}} \path \bot, 
    R: \rightv{\bot_{H_0}} \path \bot \})
\end{array}\]
where $\uplus$ is the disjoint union operation on sets and $\bot$ is a new
distinguished element that is related by new paths to the distinguished
elements of $G_\bullet$ and $H_\bullet$.
\end{definition}

\begin{definition}[Smash product]
The \emph{smash product} of two pointed groupoids $G_\bullet$ and
$H_\bullet$, written $G_\bullet \smashproduct H_\bullet$ is defined as
follows:
\[\begin{array}{l}
(G_0, \bot_{G_0}, \mathit{paths}_{G_0}) \smashproduct
(H_0, \bot_{H_0}, \mathit{paths}_{H_0}) = \\
((G_0 \times H_0) \union \{ \bot \},
 \bot,
 (\mathit{paths}_{G_0} \times^p \mathit{paths}_{H_0}) \union
 \{ F_h: (\bot_{G_0},h) \path \bot \} \union 
 \{ S_g: (g,\bot_{H_0}) \path \bot \} 
\end{array}\]
where $\times$ is the cartesian product operation on sets and $\bot$ is a new
distinguished element that is related by new paths to every pair of elements
that includes either $\bot_{G_0}$ or $\bot_{H_0}$. 
\end{definition}

\begin{definition}[Denotation of types]
Each type $b$ is mapped to a \emph{pointed groupoid} $(\dt{b}, \bot_b,
\mathit{paths}_b)$ as follows:
\[\begin{array}{rcl@{\qquad}l}
\dt{0} &=& (\{ \bot_0 \}, \bot_0, \{ \mathit{loop}_0 : \bot_0 \path \bot_0 \}) &
           (\mbox{circle}) \\
\dt{1} &=& (\{ \bullet, \bot_1 \}, \bot_1, 
           \{ \mathit{loop}_1 : \bot_1 \path \bot_1 \}) &\\
\dt{(b_1 + b_2)} &=& \dt{b_1} \wedgesum \dt{b_2} & 
           (\mbox{wedge~sum}) \\
\dt{(b_1 * b_2)} &=& \dt{b_1} \smashproduct \dt{b_2} & 
           (\mbox{smash~product})
\end{array}\]
\end{definition}

Topologically $\dt{0}$ corresponds to the circle $S^1$ which has Euler
characteristic (which generalizes the notion of set cardinality) of 0. As a
groupoid, the space $\dt{0}$ includes this infinite collection of paths from
$\bot_0 \path \bot_0$:
\[\begin{array}{l}
\refl_{\bot_0} \\
\mathit{loop}_0 \\
!~\mathit{loop}_0 \\
\mathit{loop}_0 \circ \mathit{loop}_0 \\
!~(\mathit{loop}_0 \circ \mathit{loop}_0) \\
\mathit{loop}_0 \circ \mathit{loop}_0 \circ \mathit{loop}_0 \\
!~(\mathit{loop}_0 \circ \mathit{loop}_0 \circ \mathit{loop}_0) \\
\vdots
\end{array}\]
and hence, using the definition introduced by Baez and
Dolan~\cite{groupoidcard}, has groupoid cardinality 0. We will abbreviate
this collection of paths as $\mathit{loop}^n_0$ for integer $n$, where
$\mathit{loop}^0_0$ is $\refl$, and $\mathit{loop}^n_0$ abbreviates the
$n$-fold composition in one direction for positive $n$ and in the opposite
direction for negative $n$.

Note that all our pointed spaces have a similar structure: a collection of
``regular elements'' forming several connected components and a separate
isolated connected component that includes the distinguished element
$\bot$. The distinguished element $\bot$ has the infinite collection of loops
on it and, in general, may be connected to the distinguished elements of the
subspaces used in the construction. That entire cluster including $\bot$ has
Euler characteristic and groupoid cardinality 0.

We now check that the wedge sum of any of our pointed groupoids $G_\bullet$
with $\dt{0}$ is ``essentially'' $G_\bullet$ itself, i.e., that $\dt{0}$ is
the additive unit in our space of pointed groupoids. Consider:
\[\begin{array}{l}
G_\bullet \wedgesum \dt{0} = \\
(G_0, \bot_{G_0}, \mathit{paths}_{G_0}) \wedgesum 
  (\{ \bot_0 \}, \bot_0, \{ \mathit{loop}_0 : \bot_0 \path \bot_0 \}) = \\
((\leftv{G_0} \union \{ \rightv{\bot_0}, \bot \}, 
 \bot,
 (\leftv{\mathit{paths}_{G_0}} \union \{ \rightv{\mathit{loop}^n_0} , 
 \leftv{\bot_{G_0}} \path \bot, \rightv{\bot_0} \path \bot \})
\end{array}\]
One can see that this space consists of $\leftv{G_0}$ with whatever structure
is on it and another connected component consisting of $\leftv{\bot_{G_0}}$,
$\rightv{\bot_0}$, and the new $\bot$ for the entire space. This cluster has
cardinality 0 and hence the new space is ``equivalent'' to the original
$G\bullet$ as we have essentially only changed the representation of 0.

Therefore, since the wedge sum is evidently commutative and associative, it
is at least plausible that the structure $(\wedgesum, \dt{0}$ is a symmetric
monoidal category. All of that will be formalized later. The categorical
morphisms must obviously map distinguished elements to distinguished elements
and preserve the structure of the path. It is also reasonably easy to see
that the structure $(\smashproduct, \dt{1})$ is another symmetric monoidal
category. Even better one can check that for any pointed groupoid
$G_\bullet$, we have $G_\bullet \smashproduct \dt{0}$ gives us a very
complicated 0 but still a 0. I didn't check distributivity but I suspect it
holds too.

@@

@@

We have constructed a good candidate for pointed homotopy category.
